Nengo: A Python tool for building large-scale functional brain models

### 2013-10-02 nengo.tex
Edits from Emily's feedback

It is often said that neuroscience
is data-rich but theory-poor.
Although computational neuroscience
has grown significantly,
this phrase continues to be repeated
due to a lack of satisfying
theories of the brain.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail,
but behavior has not emerged
from data-driven large scale models.
Similarly, theoretical frameworks
consisting of a generic algorithm
replicated many times
(e.g., the memory-prediction framework \cite{TODO}
and the pattern recognition theory of mind \cite{TODO})
have not been shown to solve problems
that the generic algorithm alone cannot solve.

Nengo is a neural simulator
based on a theoretical framework proposed
by Eliasmith \& Anderson \citeyearpar{TODO}
called the Neural Engineering Framework
(NEF).
Nengo was recently used to implement Spaun,
currently the world's largest functional brain model
\cite{TODO}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including list memory and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to remember the list $1234$.
It then provides its answer by
sending motor commands to a simulated arm model,
tracing out digits.
If asked to recall the remember list,
it would trace out the digits $1234$.

While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules that we map
onto the brain areas hypothesized
to perform those functions.
Like data-driven large scale models,
Spaun is able to match experimental data
across the neuroscientific spectrum,
from the single cell to the behavioral level.
However, instead of constructing the model
based on data and hoping behavior emerges,
we construct the model
based on behavior
and see that the data emerges.
This is made possible by a theoretical framework
like the memory-prediction framework \cite{TODO},
but unlike it and other frameworks,
the NEF enables Spaun to employ
many different algorithms
based on the behavior being performed,
rather than defining a single algorithm
and hypothesizing that replicating it
many times will enable these behaviors.

@@

JavaNengo is a suite of tools providing
a neural simulator and graphical user interface (GUI)
implemented in Java,
and a scripting interface implemented in Jython
(see Table~\ref{tab:nengo-vers} for more detail).
JavaNengo's simulator was created
with the intention of being a general
neural simulator that included
the methods of the NEF
as one of several
methods for creating neural models.
As a result, its implementation
of important NEF components
are nested in deep class hierarchies,
adding unnecessary layers of complexity
for developers wishing
to extend those NEF components.
As a representative example,
the most often used class for making
NEF models, \texttt{NEFEnsemble},
extends from five ancestor classes
and implements seven interfaces.
Adding support for synaptic plasticity
in \texttt{NEFEnsemble}
involved adding or significantly modifying
25 Java classes \cite{TODO}.

The JavaNengo GUI provided
a drag-and-drop interface in which to create NEF models.
However, as it was developed
before the scripting interface,
the GUI manipulates simulator objects directly.
Typically, modellers start by following
tutorials that use the GUI interface,
and transition to the scripting interface
in order to implement more complicated models.
Since the GUI modifies simulator objects directly,
modelling concepts learned in the GUI
have to be relearned in the scripting interface.

@@

Finally, perhaps the most significant limitation
of the JavaNengo tools is their performance.
While Java has good support for threads,
it does not have well-supported libraries
for doing mathematical operations on large matrices quickly.
Additionally, it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which can perform the types of computations
that neural models require quickly and in parallel.

@@

PyNengo,
the subject of the remainder of this paper,
is a Python package for
defining and simulating
neural models using the NEF.
It targets the CPython interpreter,
meaning that it integrates seamlessly
with other scientific Python tools.
Arbitrary CPython programs
can use Nengo models,
opening up possibilities
for using neurally implemented algorithms
in applications and games.

PyNengo's object model is simple and minimal,
making it easy to
document, test, and modify.
It also includes a scripting interface to those objects
based on JavaNengo's scripting interface \cite{TODO},
but with several modifications
to allow for easier incorporation
of user-defined networks,
new neuron models, and learning rules.
Additionally, PyNengo's scripting interface
contains methods to record simulated data
and access that data after simulation.

Model creation and simulation are decoupled,
allowing for models to be exported
to other simulators,
such as those accessible with PyNN.
Simulators can also be written
to leverage specific platforms.
Currently, we have implemented
a platform-independent simulator
that uses NumPy for vectorized computations,
and a PyOpenCL-based
simulator that takes full advantage of GPUs
and multicore CPUs.

@@

PyNengo overcomes all of JavaNengo's limitations.
It can simulate most models
currently simulated by JavaNengo
using 11\% as many lines of code,\footnote{
  JavaNengo is 40,000 lines of code, not including the GUI,
  and PyNengo is 5000 lines of code. TODO}
and depending on
only three external libraries instead of ten.
The OpenCL simulator can simulate
a large model TODO times
faster than JavaNengo's simulator.
This, along with a codebase
that is well tested and
easy to understand and extend,
provides a platform
that can simulate larger and more complex
models than Spaun,
and can therefore further
test the NEF as a theory of the brain.

@@

\paragraph{Representation}
Information is encoded by populations of neurons.
We consider that information
to be time-varying vectors of real numbers,
in order to reason with that information
using conventional mathematics.
By doing this, we can \textit{encode}
those vectors by injecting
specific amounts of current into
single neuron models based on
the vector being encoded.
This drives the neuron,
causing it to spike.
We can then estimate
the originally encoded vector
through a \textit{decoding} process.
This idea is similar to population coding
\cite{TODO}, but generalized
to vectors of arbitrary length.

Figure~\ref{fig:nef} depicts
the encoding (A) and decoding (B) process.
In the encoding process, the input signal drives
each neuron based on its \textit{tuning curve},
which describes how much
that neuron fires for a given input signal.
The tuning curve summarizes the gain
of a neuron (how quickly the activity rises),
the bias (the activity of a neuron given no signal),
and the encoding weight
(the part of the vector space
that causes the neuron to fire most strongly).
This results in a predictable pattern of firing
given the input signal.

@@

\paragraph{Transformation}
Neurons communicate through
unidirectional connections called synapses.
When a neuron spikes,
it releases neurotransmitter across the synapse,
which causes some amount of current
to be imparted in the downstream neuron.
Many factors affect the
amplitude of the current imparted;
we summarize those factors
in a scalar connection weight
representing the strength
of the connection between two neurons.
In order to compute any function,
we can set the connection weights between
two populations to be the product of
the decoding weights for that function
for neurons in the first population
and the encoding weights
for the downstream neuron.

@@

\paragraph{Nengo}
Large models can be built
by thinking of the principles of the NEF
as building blocks that can be put together
to describe neural systems.
The goal of Nengo is to allow
modellers to describe neural systems
in terms of what information is being represented,
how that information is transformed,
and what dynamics are required
to perform cognitively relevant
functions with biologically constrained
neural components.
Those descriptions are then
translated to a network
of interconnected neurons.
This situates Nengo
as a ``neural compiler''
that translates
a high-level functional model
to a low-level neural model
that implements the high-level function.

@@

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
An ensemble must be given as arguments
its dimensionality
(i.e., the length of the vector it represents),
and an object that describes
the population of neurons.
For example, \texttt{nengo.LIF(50, tau\_ref=0.002)}
describes a population
of 50 leaky integrate-and-fire neurons \cite{TODO} % LIF cite
with a refractory time of 0.002 seconds.
The \texttt{nengo.LIF} class defines
the parameters of the LIF neurons symbolically
so that each simulator can compute
the LIF nonlinearity efficiently.
The neuron model used by the ensemble
is changed by passing in a different symbolic neuron object;
however, the simulator used must be aware
of that neuron model.

@@

>>>>

It is often said that neuroscience
is data-rich but theory-poor.
Although computational neuroscience
has grown significantly,
this phrase continues to be repeated
due to a lack of satisfying
theories of the brain.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail
with the thought that
cognition will emerge by connecting
detailed neuron models according
to the statistics of biological connections \cite{TODO}; % blue brain
cognition has not yet emerged
from data-driven large scale models.
Similarly, theoretical frameworks
that use a generic algorithm
replicated many times % reword based on papers below
are unable to solve more complex problems
than the generic algorithm alone
(e.g., \cite{TODO} \cite{TODO}). % Hawkins, Kurzweil

Nengo is a neural simulator
based on a theoretical framework proposed
by Eliasmith \& Anderson \citeyearpar{TODO}
called the Neural Engineering Framework
(NEF).
Nengo was recently used to implement Spaun,
currently the world's largest functional brain model
\cite{TODO}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including memorizing lists and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to remember the list $1234$.
It then provides its answer by
sending motor commands to a simulated arm model,
tracing out digits.
If asked to recall the remembered list,
it would trace out the digits $1234$.

While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules that we map
onto the brain areas hypothesized
to perform those functions.
Like data-driven large scale models,
Spaun is able to match experimental data
across the neuroscientific spectrum,
from the single cell to the behavioral level.
However, instead of constructing the model
based on data and hoping behavior emerges,
we construct the model
based on behavior
and see that the data emerges.
This is made possible by a theoretical framework
like the memory-prediction framework \cite{TODO},
but unlike it and other frameworks,
the NEF enables Spaun to employ
different sets of algorithms
depending on the behavioral task performed,
rather than defining a single algorithm
and hypothesizing that replicating it
many times will enable these behaviors.

@@

JavaNengo is a suite of tools providing
a neural simulator and graphical user interface (GUI)
implemented in Java,
and a scripting interface implemented in Jython
(as detailed in Table~\ref{tab:nengo-vers}).
JavaNengo's simulator was created
with the intention of being a general
neural simulator that included
the methods of the NEF
as one of several
methods for creating neural models.
As a result, its implementation
of important NEF components
are nested in deep class hierarchies,
adding unnecessary layers of complexity
for developers wishing
to extend those NEF components.
As a representative example,
the most often used class for making
NEF models, \texttt{NEFEnsemble},
extends from five ancestor classes
and implements seven interfaces.
Adding support for synaptic plasticity
in \texttt{NEFEnsemble}
involved adding or significantly modifying
25 Java classes \cite{TODO}.

The JavaNengo GUI provided
a drag-and-drop interface in which to create NEF models.
However, as it was developed
before the scripting interface,
the GUI manipulates simulator objects directly,
rather than constructing a script
that manipulates those simulator objects.
Typically, modellers start by following
tutorials that use the GUI interface,
and transition to the scripting interface
in order to implement more complicated models;
since the GUI modifies simulator objects directly,
modelling concepts learned in the GUI
have to be relearned in the scripting interface.

@@

Finally, perhaps the most significant limitation
of the JavaNengo tools is performance.
While Java has good support for threads,
it does not have well-supported libraries
for doing mathematical operations on large matrices quickly.
Additionally, it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which can perform the types of computations
that neural models require quickly and in parallel.

@@

PyNengo is a Python package for
defining and simulating
neural models using the NEF.
It targets the CPython interpreter,
meaning that it integrates seamlessly
with other scientific Python tools.
Arbitrary CPython programs
can use Nengo models,
opening up possibilities
for using neurally implemented algorithms
in applications and games.

PyNengo has a simple object model,
which makes it easy to
document, test, and modify.
PyNengo's scripting interface to those objects
is modified from JavaNengo's scripting interface \cite{TODO}
to easily incorporate user-defined networks,
new neuron models and learning rules,
and multiple simulators.
Additionally, the scripting interface
contains functions to record data during a simulation,
and access that data after the simulation.

Model creation and simulation are decoupled,
allowing for models to be exported
to other simulators,
such as those accessible with PyNN.
Simulators can also be written
to leverage special-purpose hardware.
Currently, we have implemented
a platform-independent simulator
that uses NumPy for vectorized computations,
and a PyOpenCL-based
simulator that takes full advantage of GPUs
and multicore CPUs.

@@

PyNengo overcomes JavaNengo's limitations.
It can simulate most models
currently simulated by JavaNengo
using 11\% as many lines of code,\footnote{
  JavaNengo is 40,000 lines of code, not including the GUI,
  and PyNengo is 5000 lines of code. TODO}
and depends on
only three external libraries instead of ten.
The OpenCL simulator can simulate
a large model TODO times
faster than JavaNengo's simulator.
This, along with a codebase
that is well tested and
easy to understand and extend,
provides a platform
that can simulate larger and more complex
models than Spaun,
and can therefore further
test the NEF as a theory of the brain.

@@

\paragraph{Representation}
Information is encoded by populations of neurons.
We represent information
with time-varying vectors of real numbers
in order to manipulate that information
using conventional mathematics.
By doing this, we can \textit{encode}
those vectors by injecting
specific amounts of current into
single neuron models based on
the vector being encoded.
This drives the neuron,
causing it to spike.
We can then estimate
the originally encoded vector
through a \textit{decoding} process.
This idea is similar to population coding
\cite{TODO}, but generalized
to vectors of arbitrary length.

Figure~\ref{fig:nef} depicts
the encoding (A) and decoding (B) process.
In the encoding process, the input signal drives
each neuron based on its \textit{tuning curve},
which describes how likely
that neuron is to fire for a given input signal.
The tuning curve summarizes the gain
of a neuron (how quickly the activity rises),
the bias (the activity of a neuron given no signal),
and the encoding weight
(the part of the vector space
that causes the neuron to be the most active).
This results in a predictable pattern of firing
given the input signal.

@@

\paragraph{Transformation}
Neurons communicate through
unidirectional connections called synapses.
When a neuron spikes,
it releases neurotransmitter across the synapse,
which causes some amount of current
to be imparted in the downstream neuron.
Many factors affect the
amplitude of the current imparted;
we summarize those factors
in a scalar connection weight
representing the strength
of the connection between two neurons.
In order to compute any function,
we set the connection weights between
two populations to be the product of
the decoding weights for that function
for neurons in the first population
and the encoding weights
for the downstream neuron.

@@

\paragraph{Nengo}
Large models can be built
by using of the principles of the NEF
as connectable components
that describe neural systems,
just as a circuit diagram
describes an electronic circuit.
The goal of Nengo is to allow
modellers to describe neural systems
in terms of what information is being represented,
how that information is transformed,
and what dynamics are required
to perform cognitively relevant
functions with biologically constrained
neural components.
Those descriptions are then
translated to a network
of interconnected neurons.
This situates Nengo
as a ``neural compiler''
that translates
a high-level functional model
to a low-level neural model
that implements the high-level function.

@@

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
An ensemble must be given as arguments
its dimensionality
(i.e., the length of the vector it represents),
and an object that describes
the population of neurons.
For example, \texttt{nengo.LIF(50, tau\_ref=0.002)}
describes a population
of 50 leaky integrate-and-fire neurons \cite{TODO} % LIF cite
with a refractory time of 2 milliseconds.
The \texttt{nengo.LIF} class defines
the parameters of the LIF neurons symbolically
so that each simulator can compute
the LIF nonlinearity efficiently.
The neuron model used by the ensemble
is changed by passing in a different symbolic neuron object;
however, the simulator used must be aware
of that neuron model.

@@

% Also acknowledge Bryan for making JavaNengo, and edits
